Two years of explicit CiTO annotations

### 2023-01-09 article.Rmd
Applied suggestions by Rajarshi

Citations are an essential aspect of research communication and have become the basis of many
metrics in the academic world. Some see citation counts as a mark of scientific impact or
even quality, but in reality the reasons for citing other work are manifold which makes
the interpretation more complicated than a single citation count can reflect.
Two years ago, the *Journal of Cheminformatics* proposed the CiTO Pilot for the adoption of
a practice of annotating citations with their citation intentions.
Basically, when you cite a journal article or dataset (or any other source), you also explain
why specifically you cite that source. Particularly,
the agreement and disagreement and reuse of methods and data are of interest.
This article explores what happened after the launch of the pilot. We summarize how
authors in the *Journal of Cheminformatics* used the pilot, shows citation annotations
are distributed with Wikidata, visualized with Scholia, discusses adoption outside BMC,
and ponders about what needs to happen next.

@@

With the use of citations increasingly being picked up to help researchers with tools like
scite.ai&nbsp;[@citesAsAuthority:Nicholson_2021] en Connected Papers&nbsp;[@citesAsAuthority:ConnectedPapers], having typed
citations will help us explore literature. Therefore, the *Journal of Cheminformatics*
started a pilot for the adoption of annotating citations with their annotations&nbsp;[@discusses:Willighagen_2020].

@@

# Adoption by *Journal of Cheminformatics* authors

@@

# Technological innovation

To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support:
[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html).
Here, the author indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
The manuscript can then be converted to a Microsoft Word file with Pandoc ([https://pandoc.org/](https://pandoc.org/)) for submission to the journal. For publishers it will be interesting to note that the Pandoc can be directly converted in
the *journal article tag suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].

@@

# What is next?

>>>>

Citations are an essential aspect of research communication and have become the basis of many
evaluation metrics in the academic world. Some see citation counts as a mark of scientific impact or
even quality, but in reality the reasons for citing other work are manifold which makes
the interpretation more complicated than a single citation count can reflect.
Two years ago, the *Journal of Cheminformatics* proposed the CiTO Pilot for the adoption of
a practice of annotating citations with their citation intentions.
Basically, when you cite a journal article or dataset (or any other source), you also explain
why specifically you cite that source. Particularly,
the agreement and disagreement and reuse of methods and data are of interest.
This article explores what happened after the launch of the pilot. We summarize how
authors in the *Journal of Cheminformatics* used the pilot, shows citation annotations
are distributed with Wikidata, visualized with Scholia, discusses adoption outside BMC,
and finally present some thoughts on what needs to happen next.

@@

With the use of citations increasingly being picked up to help researchers with tools like
scite.ai&nbsp;[@citesAsAuthority:Nicholson_2021] en Connected Papers&nbsp;[@citesAsAuthority:ConnectedPapers], having typed
citations will help us explore literature. Therefore, the *Journal of Cheminformatics*
started a pilot to explore capturing the intent of citations using annotations&nbsp;[@discusses:Willighagen_2020].

@@

# Adoption by *Journal of Cheminformatics* Authors

@@

# Technological Innovation

To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support:
[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html).
Here, the author indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
The manuscript can then be converted to a Microsoft Word file with Pandoc ([https://pandoc.org/](https://pandoc.org/)) for submission to the journal. For publishers it will be interesting to note that the Pandoc can be directly converted in
the *Journal Article Tag Suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].

@@

# What is Next?
